---
title: "Assignment1"
author: "Tzu Yu Huang"
date: "2022-12-08"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```


## 1. Set up and read in the data

```{r}
library(dplyr)
library(lmtest)
master <- read.csv('Assignment1/master_data.csv')
lab <- read.csv('Assignment1/t016.csv')
adv <- read.csv('Assignment1/t05xc.csv')
```

## 2. Take a look and organize the data.

According to the instructions, I'll acquire the patients treatment assignment and region information from dataset "master", and potassium value information from the datasets "lab" and "adv".  

There are too many variables in the master dataset, so I'll just extract those that are needed. 

```{r}
mas <- master[, c("master_id", "treat", "region")]
```

Take a look at how many patients are there and how many different region and treatment assignments.

```{r}
length(mas$master_id)
length(unique(mas$master_id))
table(mas$region)
table(mas$treat)
```

So we can see that there are two treatment groups, two regions, and no duplicated master_id in dataset "mas".

```{r}
head(lab)
head(adv)
```

## 3. Discover patients that have experienced Hyperkalemia before (defined as any potassium value >=5.5)

First acquire a list from dataset "lab"

```{r}
lab_list <- lab[lab$k >= 5.5 & !is.na(lab$k), ]$master_id
```

Next from dataset "adv"

```{r}
adv_list <- adv[adv$k >= 5.5 & !is.na(adv$k), ]$master_id
```

Then combine the list to get patients that have experienced hyperkalemia before.

```{r}
list <- c(lab_list, adv_list) %>% unique()
```

Combine these information with the dataset "mas". "hyperk" = 1 indicates this patient has experienced hyperkalemia before.

```{r}
mas$hyperk <- ifelse(mas$master_id %in% list, 1, 0 ) 
```

Now that we have a dataset that has information about treatment, hyperkalemia, region, and patient identifier. And No NA value or unusually small or large value being observed.

```{r}
summary(mas)
```

## 4. Answer the questions

Since the outcome of interest for this question is binary by nature (whether or not a patient has ever experienced hyperkalemia), I will implement a logistic regression model for the following questions. 

## a) Is there evidence that the drug is associated with hyperkalemia?

The first model I fit only have treatment assignmnet as independent variable.

```{r}
mas$treat <- as.factor(mas$treat)
mas$region <- as.factor(mas$region)
logit <- glm(hyperk ~ treat, data = mas, family = "binomial" (link = 'logit'))
summary(logit)
```

And yes, there is evidence that the drug is associated with hyperkalemia, the treatment is statistically significantly associated with hyperkalemia (p value = 1.47e-15).

Next, I fit another model while controlling for the main effect of geographic region. 

```{r}
reglogit <- glm(hyperk ~ treat + region, data = mas, family = "binomial" (link = 'logit'))
summary(reglogit)
```

```{r}
lrtest(reglogit, logit)
```

The treatment is still statistically significantly associated with hyperkalemia after adjusting for geographic region (p value = 1.27e-15). And the new model with region also shows a statistically significant better fit compare to model without region. 

## b) Does the drug effect depend on geographic region?

To see if the effect of drug depend on geographic region, I included an interaction term.

```{r}
intlogit <- glm(hyperk ~ treat*region, data = mas, family = "binomial"(link = 'logit'))
summary(intlogit)
```

```{r}
lrtest(intlogit, reglogit)
lrtest(intlogit, logit)
```

It appears that the effect of drug does depend on geographic region (p value = 4.16e-06). Interestingly, geographical region itself no longer has a statistically significant relationship with hyperkalemia (p value = 0.703). This model with interaction term provides a lower AIC (2667.3) and statistically better fit compare to the previous two model with and without region (LR test: p value = 3.952e-06, 6.195e-12, respectively). 



