---
title: "Assignment3"
author: "Tzu Yu Huang"
date: "2022-12-08"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## 1. Setup and read in the data.

```{r}
# load the required package
library(stringr)
library(dplyr)
library(lubridate)
library(xlsx)    

# read in the file as "CCU" and inspect, understand the dataset 
CCU <- read.csv('Assignment3/202210-EDW-CCU_assessment.csv')
CCU$X <- NULL
```

## 2. Data Analysis: first entry and first exit of ICU

My general strategy is to identify rows in the dataset where the first ICU admission happened, and then the the rows where the ICU discharge happened, then use these two columns to calculate the length of stay (LOS).

Because we need to evaluate the chronological order of both admission and discharge of ICU, I start by identifying all rows of admission and discharge of ICU through text mining and "ifelse" statement, so that we can compare them in the future. 

For entering of the ICU "ENTER_ICU", there are two possible scenarios.
  1. Patients being directly admitted to the ICU per encounter
  2. Patients being transferred to the ICU per encounter

```{r}
CCU$ENTER_ICU <- 
  ifelse(
    # directly admitted to the ICU
    CCU$ADT_EVNT_NM == "ADMISSION" & grepl("ICU", CCU$FROM_DEPT, fixed = TRUE), CCU$ADT_DTTM_hidden,
    
    # being transferred to the ICU from non-ICU department
      ifelse(CCU$ADT_EVNT_NM == "TRANSFER OUT" &  grepl("ICU", CCU$FROM_DEPT, fixed = TRUE) == FALSE &  grepl("ICU", CCU$TO_DEPT, fixed = TRUE) , CCU$ADT_DTTM_hidden, NA)
         )

head(CCU)
```

Now we get the information on every time the patient entered the ICU.

Next, I tried to get the information on every time the patient left the ICU. Three possible scenarios.
  1. Transfer from an ICU department to an non-ICU department
  2. Discharge from an ICU department
  3. Still in the ICU after the date “10/3/2022"

```{r}
# The first two scenarios
CCU$OUT_ICU <- ifelse(
  # being transferred out
  CCU$ADT_EVNT_NM == "TRANSFER OUT" & grepl("ICU", CCU$FROM_DEPT, fixed = TRUE) & grepl("ICU", CCU$TO_DEPT, fixed = TRUE)== FALSE, CCU$ADT_DTTM_hidden, 
    # discharge from ICU
    ifelse(CCU$ADT_EVNT_NM == "DISCHARGE" &  grepl("ICU", CCU$FROM_DEPT, fixed = TRUE) , CCU$ADT_DTTM_hidden,  NA )
                      )

# Third scenario, still in the ICU after “10/3/2022"
stillinICU <- CCU %>% 
  group_by(ENC_NO_hidden) %>%
  filter(
    # identified by the last record per encounter indicating still in the ICU, and earlier than “10/3/2022"
    row_number()==n() &  grepl("ICU", TO_DEPT, fixed = TRUE) &  strptime( ADT_DTTM_hidden, format="%m/%d/%y %H:%M" ) < strptime( "10/3/22", format="%m/%d/%y" )
         )  %>% ungroup()

# Making the date "ADT_DTTM_hidden" of the third scenario the same as "OUT_ICU" in the main dataset "CCU"
CCU$OUT_ICU[match(stillinICU$ADT_DTTM_hidden, CCU$ADT_DTTM_hidden)] <- CCU$ADT_DTTM_hidden[match(stillinICU$ADT_DTTM_hidden, CCU$ADT_DTTM_hidden)]

head(CCU)
```

Now we've identified the record every time a patient entered and left the ICU. Since we are only interested in the first admission and discharge per encounter, I'll acquire those information by calculating the earliest (minimum) date per encounter.

```{r}
CCU <- CCU %>% 
  group_by(ENC_NO_hidden) %>%
  mutate(
    FIRST_ICU_ADMISSION = min(strptime(ENTER_ICU, format="%m/%d/%y %H:%M"), na.rm = T), 
    FIRST_ICU_DISCHARGE = min(strptime(OUT_ICU, format="%m/%d/%y %H:%M"), na.rm = T)
  ) %>%
  ungroup()
```

## 3. Data Analysis: re-entery of ICU

Now that "FIRST_ICU_ADMISSION" is being establish, but for "FIRST_ICU_DISCHARGE", I need to consider the possibility of reentering the ICU within an hour.

```{r}
# calculate the time difference between every time a patient entered the ICU "ENTER_ICU" and the earliest time they left the ICU "FIRST_ICU_DISCHARGE".
diff <- difftime(  strptime( CCU$ENTER_ICU, format="%m/%d/%y %H:%M" ), CCU$FIRST_ICU_DISCHARGE, units = "hours")

#diff >0 indicates reenter after the first discharge, diff <1 indicates the reenter is within 1 hour of the first discharge.
CCU$more <- ifelse( diff < 1 & diff >0, 1, 0 )

# record encounter identifier that has "more" equals 1 (indicating enter the ICU at least the second time)
list_more <- CCU[CCU$more == 1 & !is.na(CCU$more), ]$ENC_NO_hidden %>% unique()

# calculate the second earliest discharge date for those recorded in "list_more" 
returnICU <- 
  CCU[CCU$ENC_NO_hidden %in% list_more, ] %>%
  filter( strptime( OUT_ICU, format="%m/%d/%y %H:%M" ) != FIRST_ICU_DISCHARGE ) %>%
  group_by(ENC_NO_hidden) %>%
  mutate( 
    SECOND_ICU_DISCHARGE = min(strptime( OUT_ICU, format="%m/%d/%y %H:%M" ), na.rm = T)
  ) %>%
  ungroup()

# replacing "FIRST_ICU_DISCHARGE" from the main dataset "CCU" with the acquired "SECOND_ICU_DISCHARGE" from dataset "returnICU". Encounter identifier "ENC_NO_hidden" used as leverage.
CCU[CCU$ENC_NO_hidden %in% returnICU$ENC_NO_hidden, ]$FIRST_ICU_DISCHARGE <- returnICU$SECOND_ICU_DISCHARGE[match(CCU[CCU$ENC_NO_hidden %in% list_more, ]$ENC_NO_hidden, returnICU$ENC_NO_hidden )]
```

Now that the first discharged date has been updated to include patients that reenter the ICU within an hour at most once. I wish to repeat the process until all first discharged date has been updated, regardless of times of reentering with an hour.

```{r}
diff2 <- difftime(  strptime( CCU$ENTER_ICU, format="%m/%d/%y %H:%M" ), CCU$FIRST_ICU_DISCHARGE, units = "hours")
CCU$more2 <- ifelse( diff2 < 1 & diff2 >0, 1, 0 )
sum(CCU$more2 == 1, na.rm = T)
```

Since there are only three encounters left, I'll just update those manually.

```{r}
CCU[CCU$ENC_NO_hidden %in% CCU[CCU$more2 == 1 & !is.na(CCU$more2), ]$ENC_NO_hidden, ]

CCU[CCU$ENC_NO_hidden == 29195678735, ]$FIRST_ICU_DISCHARGE <- strptime( "9/20/21 16:35", format="%m/%d/%y %H:%M" )
CCU[CCU$ENC_NO_hidden == 29190725868, ]$FIRST_ICU_DISCHARGE <- strptime( "4/23/21 9:36", format="%m/%d/%y %H:%M" )
CCU[CCU$ENC_NO_hidden == 29188010087, ]$FIRST_ICU_DISCHARGE <- strptime( "3/26/21 14:14", format="%m/%d/%y %H:%M" )
```

## 4. Condense and Organize output dataset

After the data analysis, we have all the original data and the two required new variables,  next I'll remove the unnecessary data and calculate the length of stay in terms of hours. 

```{r}
output<- unique(CCU[, c("MRN_hidden", "ENC_NO_hidden", "FIRST_ICU_ADMISSION", "FIRST_ICU_DISCHARGE")])

output$FIRST_ICU_LOS <- difftime( output$FIRST_ICU_DISCHARGE, output$FIRST_ICU_ADMISSION, units = "hours") %>% floor() %>% as.numeric()

colnames(output)[1:2] <- c("MRN", "ENC_NO")
```

Double check if my results agree with those in file "Template".

```{r}
output[output$ENC_NO %in% c(29204845127, 29188820095, 29192933522), ]
```

It matches. Write a xlsx file with the output file.

```{r}
write.xlsx(output, 'Assignment3_Result.xlsx')
```

A few things I noticed is that, interestingly, some encounters seems to be missing the required data. For example, the first ICU admission cannot be established with encounter number 29203421031. 



